An Interpretable Machine Learning Framework for Modeling High-Resolution Spectroscopic Data\footnote{Open source code at \url{https://github.com/gully/blase}}

### 2022-08-12 paper/paper1/ms.tex
Title brainstorm, conclusions, feedback by @astrocaroline, yay!

\begin{document}
\shorttitle{Transfer learning with blas\'e}
\shortauthors{Gully \& Morley}
\title{Precision line-by-line forward models for \'echelle spectra\footnote{Open source code at \url{https://github.com/gully/blase}}}

@@

\emph{``A good enough substrate''}-- The imperfections in pre-computed stellar and substellar models have hampered their use in high-fidelity applications, where model flaws overwhelm the sigal-to-noise in the data.  The adaptability of blas\'e can be viewed as increasing the number of applications that could conceivably benefit from the rich-albeit-imperfect information enoded in these model grid spectra.  In the next section we illustrate how these adaptable models can serve as a subtrate for applications in which the imperfections in precomputed models would have otherwise precluded their use.

@@

This spirit resembles \texttt{ROBOSPECT} \citep{2013PASP..125.1164W}, and other approaches aimed at automatically measuring equivalent widths for a large number of stars.  One missing piece here is either to post-process the blas\'e-output EW catalog with chemical labels (C, N, O, Si, Ba, Eu, etc.), or incorporate that information directly into the structure of blas\'e.  The latter case is intriguing.  For example, PHOENIX may not yet be aware of the existence of some obscure Lanthanum (La) line, possibly because the line was not yet understood and incorporated at the time of PHOENIX computation.  Rather than wait for PHOENIX to receive the necessary update, blas\'e could instantiate one or more such lines in the locations guided by the atomic line list.  This line could get initialized with some small amplitude, and would get added onto the underlying PHOENIX source as yet another freely tunable line.  In this way, all the other nearby lines would be used for assigning continuum and blending, but this free-standing Lanthanum line would acquire a value properly taking into account the impact of its neighbors.

\subsection{Doppler Imaging}
The fixed $v\sin{i}$ approximation breaks down for stars with large-scale surface features.  Doppler imaging attempts to reverse engineer the surface map from the extent to which observed line profiles depart from a pristine rotational broadening kernel.  This reverse engineering step suffers from a vast number of geometrical degeneracies, but still provides useful constraints on stellar surfaces \citet{2021arXiv211006271L}.  We emphasize a distinction between A) longitudinally symmetric surface features, and B) longitudinally asymmetric surface features.  Most radial velocity practitioners think about the latter, since longitudinally asymmetric surface features imbue changing-in-time skewness to the line profiles, causing radial velocity perturbations easily detectable in radial velocity time series.  These confound exoplanet searches.

@@

Relatedly, magnetic fields information could be incorporated into blas\'e as line-by-line Land\'e $g-$factors, to allow magnetic sensitive lines to exhibit either Zeeman broadening or Zeeman splitting.  This labelling could be conducted in the same way as described in the Abundances subsection \citep{secAbundance}.

@@

\section{Conclusions}
We have introduced a line-by-line approach to forward modelling stellar, substellar, and telluric spectroscopic data.  The approach relies on a key enabling technology, automatic differentiation, that allows nearly unlimited number of spectral lines to be forward modelled simultaneously.  We initialize these lines to match precomputed synthetic stellar spectra, achieving excellent performance, and lending some confidence that the approach has a capacity to capture a tremendous amount of information at once.

>>>>

\begin{document}
\shorttitle{Transfer learning with blas\'e}
\shortauthors{Gully \& Morley}
\title{An Interpretable Machine Learning Framework for Modeling High-Resolution Spectroscopic Data\footnote{Open source code at \url{https://github.com/gully/blase}}}

@@

\emph{``A good enough substrate''}-- The imperfections in pre-computed stellar and substellar models have hampered their use in high-fidelity applications, where model flaws overwhelm the sigal-to-noise in the data.  The adaptability of blas\'e can be viewed as increasing the number of applications that could conceivably benefit from the rich-albeit-imperfect information encoded in these model grid spectra.  In the next section we illustrate how these adaptable models can serve as a subtrate for applications in which the imperfections in precomputed models would have otherwise precluded their use.

@@

\subsection{Identifying missing lines}\label{secMissingLines}
The automatic abundance determination spirit of blas\'e resembles \texttt{ROBOSPECT} \citep{2013PASP..125.1164W}, and other approaches aimed at automatically measuring equivalent widths for a large number of stars.  One missing piece here is either to post-process the blas\'e-output EW catalog with chemical labels (C, N, O, Si, Ba, Eu, etc.), or incorporate that information directly into the structure of blas\'e.  The latter case is intriguing.  For example, PHOENIX may not yet be aware of the existence of some obscure Lanthanum (La) line, possibly because the line was not yet understood and incorporated at the time of PHOENIX computation.  Rather than wait for PHOENIX to receive the necessary update, blas\'e could instantiate one or more such lines in the locations guided by the atomic line list.  This line could get initialized with some small amplitude, and would get added onto the underlying PHOENIX source as yet another freely tunable line.  In this way, all the other nearby lines would be used for assigning continuum and blending, but this free-standing Lanthanum line would acquire a value properly taking into account the impact of its neighbors.

\subsection{Doppler Imaging}
The fixed $v\sin{i}$ approximation breaks down for stars with large-scale surface features.  Doppler imaging attempts to reverse engineer the surface map from the extent to which observed line profiles depart from a pristine rotational broadening kernel.  This reverse engineering step suffers from a vast number of geometrical degeneracies, but still provides useful constraints on stellar surfaces \citep{2021arXiv211006271L}.  We emphasize a distinction between A) longitudinally symmetric surface features, and B) longitudinally asymmetric surface features.  Most radial velocity practitioners think about the latter, since longitudinally asymmetric surface features imbue changing-in-time skewness to the line profiles, causing radial velocity perturbations easily detectable in radial velocity time series.  These confound exoplanet searches.

@@

Relatedly, magnetic fields information could be incorporated into blas\'e as line-by-line Land\'e $g-$factors, to allow magnetic sensitive lines to exhibit either Zeeman broadening or Zeeman splitting.  This labelling could be conducted in the same way as described in Section \citep{secMissingLines}.

@@

\section{Conclusions}
We have introduced an interpretable machine learning approach to forward modelling stellar, substellar, and telluric spectroscopic data.  The line-by-line approach approach relies on a key enabling technology, automatic differentiation, that allows a nearly unlimited number of spectral lines to be forward modelled simultaneously.  We initialize these lines to match precomputed synthetic stellar spectra, achieving excellent performance, and lending some confidence that the approach has a capacity to capture a tremendous amount of information at once.

We demo the framework on two sources: the K5 exoplanet host star WASP 69 using a precomputed PHOENIX model, and the T6 ultracool dwarf 2MASS J08173001$-$6155158 using a precomputed Sonora-Bobcat model.  We discuss how blas\'e can be used to measure Equivalent Widths for thousands of lines automatically, understand line lists, measure rotation rates, generate surrogate models, and construct semi-empirical models.  This tool could readily have applications across stellar and substellar astronomy, including for PRV work, stellar compositions, Doppler imaging, and stellar activity.
